\LaTeXe\ guide for the Journal of Functional Programming

### 2019-10-20 papers/jfp/5-rebuilders.tex
Polish S5-7.

\subsubsection{Trace Representation}\label{sec-trace-representation}

@@

\subsubsection{Verifying Step Traces}\label{sec-step-traces}

The \Shake build system and the associated paper -- see~\S2.3.3
in~Mitchell~\shortcite{mitchell2012shake} -- use a slightly different trace
structure, called \emph{verifying step traces},
which stores less data than verifying traces, and has slightly different early cutoff semantics. Rather than
storing the \hs{Hash}~\hs{v} for each dependency, it instead stores \hs{built}
time and \hs{changed} time for each \hs{key}, and a list of dependency keys
(without the hashes). The resulting \hs{StepTrace} type resembles:

@@

This optimisation is useful, at least in the case of \Shake, to save space. A
typical cryptographic hash takes up 32 bytes, while a key (in \Shake) is an
\hs{Int} taking only~4~bytes. Furthermore, \Shake permits values to be
arbitrarily large, and permits a custom equality (two values can be bit-for-bit
unequal but considered equal by \Shake), meaning that a \hs{Hash} is not a valid
encoding. For build systems that use \hs{Applicative} dependencies, the
\hs{depends} could be omitted entirely, making the size of a \hs{StepTrace}
$O(1)$ instead of $O(n)$, where $n$ is the number of dependencies.

@@

A verifying trace deliberately records only hashes, so that it can be small. In
contrast, a \emph{constructive} trace also stores the resulting value.
Concretely, it records a list of \hs{Trace}~\hs{k}~\hs{v}~\hs{v}, where
\hs{Trace} is as defined in~\S\ref{sec-trace-representation}. Once we are
storing the complete result it makes sense to record many constructive traces
per key, and to share them with other users, providing cloud-build
functionality. We can represent this additional information with the operations:

@@

\noindent
The function \hs{recordCT} is similar to \hs{recordVT}
from~\S\ref{sec-verifying-traces}, but instead of just passing
the hash of the resulting value, we pass the actual value. The \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
current value as \emph{input}, returns a list of possible values as
\emph{output} -- there may be more than one, because some build tools are
non-deterministic, see~\S\ref{sec-non-determinism}.

@@

Any \hs{Applicative} build system using constructive traces, e.g.
\CloudBuild~(\S\ref{sec-implementation-cloud}), can index directly from the
key and the hashes of its dependencies to the output result, i.e.
\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
are stored on a central server, the client can compute the key and the hashes of
its dependencies locally, and then make a single call to the server to retrieve
the result.

@@

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$. By picking values of $n$ in between we would regain some early cutoff, at the
cost of losing such simple shallow builds, while still requiring deterministic tasks.

>>>>

\vspace{-1mm}
\subsubsection{Trace Representation}\label{sec-trace-representation}

@@

\vspace{-1mm}
\subsubsection{Verifying Step Traces}\label{sec-step-traces}

The \Shake build system and the associated paper -- see~\S2.3.3
in~Mitchell~\shortcite{mitchell2012shake} -- use a different trace structure,
called \emph{verifying step traces}, which stores less data than verifying
traces, and has slightly different early cutoff semantics. Rather than
storing the \hs{Hash}~\hs{v} for each dependency, it instead stores \hs{built}
time and \hs{changed} time for each \hs{key}, and a list of dependency keys
(without the hashes). The resulting \hs{StepTrace} type resembles:

@@

This optimisation is useful, at least in the case of \Shake, to save space. A
typical cryptographic hash takes up 32 bytes, while a key (in \Shake) is an
\hs{Int} taking only~4~bytes. Furthermore, \Shake permits values to be
arbitrarily large, and supports a custom value equality (two values can be
bit-for-bit unequal but considered equal by \Shake), hence \hs{Hash}~\hs{v} is
not a valid encoding. For applicative tasks, \hs{depends} can be omitted, making
the size of a \hs{StepTrace} $O(1)$ instead of $O(n)$, where $n$ is the number
of dependencies.

@@

A verifying trace records only hashes or time stamps, so that it can be small.
In contrast, a \emph{constructive} trace also stores the resulting value.
Concretely, it records a list of \hs{Trace}~\hs{k}~\hs{v}~\hs{v}, where
\hs{Trace} is as defined in~\S\ref{sec-trace-representation}. Once we are
storing the complete result it makes sense to record many constructive traces
per key, and to share them with other users, providing cloud-build
functionality. We can access this additional information with these operations:

@@

\noindent
The function \hs{recordCT} is similar to \hs{recordVT}
from~\S\ref{sec-verifying-traces}, but instead of just passing the hash of the
resulting value, we pass the actual value. The function \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
current value as \emph{input}, returns a list of possible values as
\emph{output} -- there may be more than one, because some build tools are
non-deterministic, see~\S\ref{sec-non-determinism}.

@@

Any \hs{Applicative} build system using constructive traces, e.g.
\CloudBuild~(\S\ref{sec-implementation-cloud}), can index directly from the
key and the hashes of its dependencies to the resulting value, e.g. using a
\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
are stored on a central server, the client can compute the key and the hashes of
its dependencies locally, and then make a single call to the server to retrieve
the result.

@@

Current build systems using deep constructive traces always record hashes of
terminal input keys, but the technique also works if we skip any number
of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$. By picking values of $n$ in between we would regain some early cutoff,
at the cost of losing such simple shallow builds, while still requiring
determinism.

### 2019-10-20 papers/jfp/6-implementations.tex
Polish S5-7.

\section{Build systems, concretely}\label{sec-implementations}\label{sec-design-space}

\begin{table}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\vspace{-2mm}
% \smaller
\centering
\begin{tabular}{lccc}
\hline
 & \multicolumn{3}{c}{\textbf{Scheduling algorithm}}\vspace{1.5mm} \\
\textbf{Rebuilding strategy}\gap  & \gap{}Topological\gap\S\ref{sec-topological}\gap & \gap{}Restarting\gap\S\ref{sec-restarting}\gap & \gap{}Suspending\gap\S\ref{sec-suspending}\gap    \\\hline
Dirty bit\hfill\S\ref{sec-dirty-bit}                                           & \Make       & \Excel & -              \\
Verifying traces\hfill\S\ref{sec-verifying-traces}                             & \Ninja      & -      & \Shake         \\
Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}           & \CloudBuild & \Bazel & -              \\
Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
\end{tabular}
\vspace{-2mm}
\end{table}

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler (\S\ref{sec-scheduler})
and a rebuilder (\S\ref{sec-rebuilder}).
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder. The result can be summarized in Table~\ref{tab-build-systems},
which tabulates combinations of the scheduling algorithm and the rebuilding
strategy, providing 12~possible build systems, 8~of which are inhabited
by existing build systems (we discuss these systems in~\S\ref{sec-background} and
\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
build systems. The most interesting unfilled spot in the table is suspending
constructive traces, which would provide many benefits, and which we title
\Cloud \Shake and explore further in~\S\ref{sec-implementation-cloud}.
% (as we plan on extending \Shake to occupy that spot)

\subsection{Concrete Implementations} \label{sec-concrete-impl}

@@

The scheduler calls the rebuilder, but passes it a \hs{fetch} function
that the latter calls when it needs the value of a dependent key.  This
callback returns control to the scheduler, which may in turn call the
rebuilder to bring the dependent key up to date, and so on.

@@

Note that the function \hs{dependencies} can only be applied to applicative
tasks, which restricts \Make to static dependencies, as reflected in the
type~\hs{Build}~\hs{Applicative}. Moreover, any other build system that uses
the \hs{topological} scheduler will also inherit the same restriction.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system; stores a dirty bit per key and calc chain
type Chain k = [k]
type ExcelInfo k = (k -> Bool, Chain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting dirtyBitRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on dirty bits
dirtyBitRebuilder :: Rebuilder Monad (k -> Bool) k v
dirtyBitRebuilder key value task = Task $ \fetch -> do
    isDirty <- get
    if isDirty key then run task fetch else return value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- liftChain $ go Set.empty
                          $ chain ++ [target | target `notElem` chain]
    modify $ mapInfo $ \(ir, _) -> (ir, newChain)
  where
    go :: Set k -> Chain k -> State (Store ir k v) (Chain k)
    go _    []       = return []
    go done (key:keys) = case tasks key of
      Nothing -> (key :) <$> go (Set.insert key done) keys
      Just task -> do
        store <- get
        let newTask :: Task (MonadState ir) k (Either k v)
            newTask = try $ rebuilder key (getValue key store) task
            fetch :: k -> State ir (Either k v)
            fetch k | k `Set.member` done = return $ Right (getValue k store)
                    | otherwise           = return $ Left k
        result <- liftStore (run newTask fetch) -- liftStore is in Fig. 8
        case result of
            Left dep -> go done $ dep : filter (/= dep) keys ++ [key]
            Right newValue -> do modify $ putValue key newValue
                                 (key :) <$> go (Set.insert key done) keys
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation (omitted)
liftChain :: State (Store ir k v) a -> State (Store (ir, Chain [k]) k v) a
\end{minted}
% \vspace{-2.5mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
% \vspace{-2.5mm}
\end{figure}

@@

After the \hs{newTask} is executed (with the help of \hs{liftStore}) there are
two cases to consider:

@@

Note that the task returned by the \hs{rebuilder} expects a total callback
function and cannot be directly executed with the partial callback \hs{fetch}.
We fix the mismatch with the function \hs{try} that relies on the standard
monad transformer \hs{ExceptT} from the \cmd{transformers} library. We also
need the helper \hs{liftChain}, whose implementation we omit since it
is analogous to \hs{liftStore} in Fig.~\ref{fig-make-implementation}.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system; stores verifying traces
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key (hash value) (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        modify $ recordVT key (hash newValue) [ (k, hash v) | (k, v) <- deps ]
        return newValue
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store =
    fst $ execState (fetch target) (store, Set.empty)
  where
    fetch :: k -> State (Store i k v, Set k) v
    fetch key = do
        done <- gets snd
        case tasks key of
            Just task | key `Set.notMember` done -> do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                newValue <- liftRun newTask fetch
                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
                return newValue
            _ -> gets (getValue key . fst) -- fetch the existing value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Run a task using a callback that operates on a larger state (omitted)
liftRun :: Task (MonadState i) k v
        -> (k -> State (Store i k v, Set k) v) -> State (Store i k v, Set k) v
\end{minted}
% \vspace{-1mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
% \vspace{-3mm}
\end{figure}

@@

As discussed in \S\ref{sec-step-traces}, \Shake actually use verifying step traces, but we
choose to focus on the more explicit verifying traces. We have implemented
verifying step traces in our framework, and they compose with schedulers as
you would hope.

@@

Our models of \Buck~\cite{buck} and \Nix~\cite{dolstra2004nix} use the rebuilder
based on deep constructive traces (\S\ref{sec-deep-constructive-traces}), called
\hs{dctRebuilder}, whose implementation we omit since it is very similar to that
of \hs{ctRebuilder}. \Buck uses the \hs{topological} scheduler and is therefore
an applicative build system, whereas \Nix uses the \hs{suspending} scheduler and
is monadic.

>>>>

\section{Build Systems, Concretely}\label{sec-implementations}\label{sec-design-space}

\begin{table}
\vspace{2mm}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\vspace{-3mm}
% \smaller
\centering
\begin{tabular}{lccc}
\hline
 & \multicolumn{3}{c}{\textbf{Scheduling algorithm}}\vspace{1.5mm} \\
\textbf{Rebuilding strategy}\gap  & \gap{}Topological\gap\S\ref{sec-topological}\gap & \gap{}Restarting\gap\S\ref{sec-restarting}\gap & \gap{}Suspending\gap\S\ref{sec-suspending}\gap    \\\hline
Dirty bit\hfill\S\ref{sec-dirty-bit}                                           & \Make       & \Excel & -              \\
Verifying traces\hfill\S\ref{sec-verifying-traces}                             & \Ninja      & -      & \Shake         \\
Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}           & \CloudBuild & \Bazel & -              \\
Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
\end{tabular}
\vspace{-2mm}
\end{table}

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler (\S\ref{sec-scheduler})
and a rebuilder (\S\ref{sec-rebuilder}).
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder. The result can be summarized in Table~\ref{tab-build-systems},
which tabulates 12 possible combinations, 8~of which are inhabited
by existing build systems (we discuss these systems in~\S\ref{sec-background}
and~\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
build systems. The most interesting unfilled spot in the table corresponds to
a suspending scheduler composed with a constructive trace rebuilder. Such a
build system would provide many benefits; we title it \Cloud \Shake and explore
further in~\S\ref{sec-implementation-cloud}.

\vspace{-1mm}
\subsection{Concrete Implementations} \label{sec-concrete-impl}

@@

The scheduler runs the \hs{Task} returned by the rebuilder passing it a
\hs{fetch}~callback~that the task uses to find values of its dependencies.
The callback returns control to \mbox{the scheduler}, which may in turn call
the rebuilder again to bring another key up to date, and so on.

@@

Note that the function \hs{dependencies} can only be applied to applicative
tasks, which restricts \Make to static dependencies, as reflected in the
type~\hs{Build}~\hs{Applicative}. Any other build system that uses
the \hs{topological} scheduler will inherit the same restriction.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system; stores a dirty bit per key and calc chain
type Chain k = [k]
type ExcelInfo k = (k -> Bool, Chain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting dirtyBitRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on dirty bits
dirtyBitRebuilder :: Rebuilder Monad (k -> Bool) k v
dirtyBitRebuilder key value task = Task $ \fetch -> do
    isDirty <- get
    if isDirty key then run task fetch else return value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- liftChain $ go Set.empty
                          $ chain ++ [target | target `notElem` chain]
    modify $ mapInfo $ \(ir, _) -> (ir, newChain)
  where
    go :: Set k -> Chain k -> State (Store ir k v) (Chain k)
    go _    []       = return []
    go done (key:keys) = case tasks key of
      Nothing -> (key :) <$> go (Set.insert key done) keys
      Just task -> do
        store <- get
        let newTask :: Task (MonadState ir) k (Either k v)
            newTask = try $ rebuilder key (getValue key store) task
            fetch :: k -> State ir (Either k v)
            fetch k | Set.member k done = return $ Right (getValue k store)
                    | otherwise         = return $ Left k
        result <- liftStore (run newTask fetch) -- liftStore is in Fig. 8
        case result of
            Left dep -> go done $ dep : filter (/= dep) keys ++ [key]
            Right newValue -> do modify $ putValue key newValue
                                 (key :) <$> go (Set.insert key done) keys
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation (omitted)
liftChain :: State (Store ir k v) a -> State (Store (ir, Chain [k]) k v) a
\end{minted}
% \vspace{-2.5mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
% \vspace{-2.5mm}
\end{figure}

@@

After the \hs{newTask} is executed (using \hs{liftStore}), there are two cases
to consider:

@@

Note that the task returned by the \hs{rebuilder} expects a total callback
function and cannot be executed with the partial callback \hs{fetch}. We fix the
mismatch with the function \hs{try} that relies on the standard monad
transformer \hs{ExceptT}~\cite{liang1995monad}. The helper \hs{liftChain} is
analogous to \hs{liftStore} in Fig.~\ref{fig-make-implementation}, so we omit
its implementation.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system; stores verifying traces
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key (hash value) (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        modify $ recordVT key (hash newValue) [ (k, hash v) | (k, v) <- deps ]
        return newValue
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store =
    fst $ execState (fetch target) (store, Set.empty)
  where
    fetch :: k -> State (Store i k v, Set k) v
    fetch key = do
        done <- gets snd
        case tasks key of
            Just task | Set.notMember key done -> do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                newValue <- liftRun newTask fetch
                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
                return newValue
            _ -> gets (getValue key . fst) -- fetch the existing value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Run a task using a callback that operates on a larger state (omitted)
liftRun :: Task (MonadState i) k v
        -> (k -> State (Store i k v, Set k) v) -> State (Store i k v, Set k) v
\end{minted}
\vspace{-1mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{-3mm}
\end{figure}

@@

As discussed in \S\ref{sec-step-traces}, \Shake actually use verifying step
traces, but here we choose to focus on the more explicit verifying traces. We
have implemented verifying step traces in our framework, and they compose with
schedulers as you would hope.

@@

Our models of \Buck~\cite{buck} and \Nix~\cite{dolstra2004nix} use the rebuilder
based on deep constructive traces (\S\ref{sec-deep-constructive-traces}), called
\hs{dctRebuilder}, whose implementation we omit since it is very similar to that
of \hs{ctRebuilder}. \Buck uses the \hs{topological} scheduler and is an
applicative build system, whereas \Nix uses the \hs{suspending} scheduler and
is therefore monadic.
